---
title: "Machine learning case study"
author: "Anna Przybyłowska, Gurbet Güngören, Wojciech Tomczak, Witold Taisner"
date: "5/4/2021"
output:
  slidy_presentation: 
    font_adjustment: -0.5
    theme: sandstone
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(massage = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(tidy = TRUE)
```

## Description of dataset

source: https://www.kaggle.com/jsphyg/weather-dataset-rattle-package

- This dataset contains information about daily weather observation in Australia over 10 years 
(2007 - 2017).

- It regards:
  - date
  - location
  - temperature (degrees celsius)
  - rainfall (mm)
  - evaporation (mm)
  - sunshine (hours)
  - wind direction and speed (km/h)
  - humidity (percent)
  - atmospheric pressure (hPa)
  - fraction of sky obscured by cloud ("oktas")
  - information if it rained that day (Yes/No, Yes if rain > 1mm)
  - information if it rained the next day (Ye/No, Yes if rain > 1mm)

- For most of the weather parameters, there is information about measurements at 9am and 3pm, for some maximal/minimal values is also given.


## Dataset
```{r dataset, echo = F, warning = F, message = F, fig.height=7}
library(tidyr)
library(ggplot2)
library(dplyr)
library(DT)

rain = read.csv("weatherAUS.csv") %>% as_tibble()
datatable(head(rain, 1000), rownames = F, class = 'cell-border stripe',
          option = list(
  scrollX = TRUE,
  pageLength = 6
))
```


## Preprocessing

Percentage of missing data

```{r plot1, fig.height=5, fig.width=7, fig.align='center', echo = F}
counting = rain %>% as_tibble() %>% select(everything()) %>%
  summarise_all(funs(sum(is.na(.))))
counting = (counting /145460) 
counting = counting %>% gather("column", "missing data", 1:23)
counting = arrange(counting, desc(`missing data`))

ggplot(counting, aes(x = reorder(column, -`missing data`), y = `missing data`)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(x = "attributes", y = "missing data [%]") + 
  theme(text = element_text(size=10))
```

## Removing missing values and detecting outliers

- We remove all missing values for categorical attributes with string values since by doing that we won't loose too much data.

```{r remove missing values}
rain = filter(rain, !is.na(rain$RainTomorrow) & !is.na(rain$RainToday) & !is.na(rain$WindGustDir)
              & !is.na(rain$WindDir9am) & !is.na(rain$WindDir3pm))

```

- For categorical attributes with string values there are no outliers/ incorrect values
  - It can be easily checked by summarizing the data (count values) or by using WEKA
- We use outliers detected by boxplot for each column, store them and then replace value of specific parameter with NA
- Boxplots use interquartile range criterion
- We will replace missing values with predicted ones

```{r outliers}
outlier_detection_data = select(rain, -c("RainToday","RainTomorrow","WindGustDir",
                                         "WindDir9am","WindDir3pm","Date", "Location"))

for(column in colnames(outlier_detection_data)){
  out = boxplot.stats(outlier_detection_data[[column]])$out
  out_id = which(rain[[column]] %in% c(out))
  for(id in out_id){
    rain[[column]][id] = NA
  }
}
```

## Correctness of data

- Some attributes must have values in specific ranges\
\
  - Clouds in [0,8]\
  \
  - Humidity in [0,100]\
  \
  - Sunshine in [0,24]\
  \
  - Rainfall, evaporation, wind speed, pressure ≥ 0\
  \
- Outlier test may have not detected them\

## Correctness of data - table

```{r table_ranges, echo = F}
tmp = do.call(
  rbind, lapply(
    select(rain, c("Cloud9am","Cloud3pm","Humidity9am", "Humidity3pm",
                       "Sunshine", "Rainfall", "Evaporation", "WindGustSpeed",
                       "WindSpeed9am", "WindSpeed3pm", "Pressure3pm", "Pressure9am")),
    summary
  )
)
tmp = data.frame(tmp)
tmp = tmp[-c(2:5,7)]
datatable(tmp, rownames = T, class = 'cell-border stripe',
  option = list(
  scrollX = TRUE,
  pageLength = 6
))

```

- We can see some incorrect values in clouds attribute, therefore we set those values to NA.
```{r outliers2}
rain$Cloud9am[!(rain$Cloud9am>=0 & rain$Cloud9am<=8)] = NA
rain$Cloud3pm[!(rain$Cloud3pm>=0 & rain$Cloud3pm<=8)] = NA

```

## Percentage of missing data

We visualize missing data one more time, after detecting outliers, checking for correctness and removing NA values for categorical attributes with string values.  

```{r plot2, echo = F, fig.height=5, fig.width=7, fig.align='center'}
counting = rain %>% as_tibble() %>% select(everything()) %>%
  summarise_all(funs(sum(is.na(.))))
counting = (counting /123710) 
counting = counting %>% gather("column", "missing data", 1:23)
counting = arrange(counting, desc(`missing data`))

ggplot(counting, aes(x = reorder(column, -`missing data`), y = `missing data`)) +
  geom_bar(stat = "identity") + 
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(x = "attributes", y = "missing data [%]") +
  theme(text = element_text(size=10))
```

## Dealing with NA's: round two

- we would loose too much data if we just dropped all missing values (more than a half)
- no need to drop the variables (usually one should drop the attribute when there are 60-70% of missing values)
- MICE library
- multiple imputation with predictive mean matching
```{r mice, warning = F, message=F, results = 'hide'}
library(mice)
imp =  mice(rain, meth='pmm', seed=123)
rain = complete(imp)
```
```{r mice2, echo = F}
datatable(head(rain, 1000), rownames = F, class = 'cell-border stripe',
          option = list(
  scrollX = TRUE,
  pageLength = 3
))
```



## Other preprocessing

- We remove location column since we want to predict if it rains tomorrow regardless of the location (apart from it's Australia)
- We transform date to season\

``` {r otherPreprocessing, echo = F}
rain = select(rain, -Location)
rain = rain %>% mutate(Season = case_when(
  substring(Date,6,7) %in% c("12","01","02") ~ "Summer",
  substring(Date,6,7) %in% c("03","04","05") ~ "Autumn",
  substring(Date,6,7) %in% c("06","07","08") ~ "Winter",
  substring(Date,6,7) %in% c("09","10","11") ~ "Spring"
)) %>% relocate(Season)
rain = select(rain, -Date)
datatable(head(rain, 1000), rownames = F, class = 'cell-border stripe',
          option = list(
  scrollX = TRUE,
  pageLength = 6
))
write.csv(rain, file = "rain.csv", row.names = FALSE)
``` 

## Correlation matrix

```{r correlation_matrix, echo = F, fig.align='center', fig.height=7, fig.width=5}
library(ggcorrplot)
rain.data = select(rain, -c("RainToday","RainTomorrow","WindGustDir",
                            "WindDir9am","WindDir3pm","Season"))
rain.cor = cor(rain.data)
ggcorrplot(rain.cor,
           ggtheme = ggplot2::theme_minimal,
           tl.cex = 10,
           tl.srt = 90,
           colors = c("#6D9EC1", "white", "#E46726"))

```
```{r}
library(ggplot2)
ggplot(rain, aes(y = Humidity3pm, x = RainTomorrow , fill = RainTomorrow)) + 
  geom_violin() + 
  theme_bw()

```


